Fast and scalable Gaussian process modeling of stellar variability \vspace{-3\baselineskip}

### 2016-10-21 paper/ms.tex
Added discussion of Matern 3/2 kernel

Stars are variable, for better or worse:  their variability provides information about the properties
of stars, but also results in noise that must be overcome in detecting and characterizing their
planetary systems.  When analyzing astronomical times series of stars, then, there are generally two 
goals:  1). to analyze the frequency spectrum and amplitude of stellar variability to beter understand stellar
properties and evolution, and/or 2). to account for correlated noise when modeling the stellar
flux variability and spectral variations.

A promising technique for addressing both of these goals is the Gaussian Process formalism 
\citep{Rasmussen2006,Gibson2012}.  The basis of this method is to assume that stellar variability
is random, Gaussian noise, but that the random variability is correlated in time, and the nature of the
correlations is completely specified by a correlation matrix.  In its simplest form, the 
correlation matrix is modeled with a time-independent autocorrelation function (although 
the time-independence can be relaxed), including a white-noise component that may vary
with time, and the autocorrelation function is described by a functional form that may be 
simply parameterized by a relatively small number of parameters, sometimes referred to as 
`hyper-parameters.'  The goal of spectral analysis is to measure the parameters which
describe the autocorrelation function, thus inferring properties of the stellar variabilty
which may be used to characterize properties of the star, such as the rotation period,
the asteroseismic frequencies, or the `flicker' \citep{Aerts2010,Noyes1984,Bastien2013}.
The goal of planet detection includes measuring the Doppler shift of the star, or measuring
the decrement of flux as a planet transits a star.  To properly account for the impact of 
stellar variability on the planet parameters requires treating correlations in the data, and 
to do so also requires evaluating the correlation matrix.

The size of the correlation matrix, $N_{elements}$ scales as the square of the length of the time-series,
$N_{element}=N_{time}^2$, so as time series grow in size, say $N_{time} =10^{5-6}$, the storage may become 
prohibitive, $N_{element} = 10^{10-12}$.  In addition, this matrix must be inverted and have
its determinant evaluated to compute the likelihood function, which takes of order $N_{operations}
= N_{time}^3 = 10^{15-18}$.  This becomes an impossible computational problem, which prohibits
using long time series datsets, such as the \texttt{Kepler} dataset $N_{time} \approx 10^{5-6}$,
\texttt{Spitzer} time series with $N_{time} \approx 10^5$, and solar time series with
$N_{time} \approx 10^7$.  So faster means of solving for the likelihood is required in these
cases.

One promising solution is to use approximate techiques based upon the fact that correlation
matrices display a great deal of symmetry.  This `hierarchical off-diagonal low-rank' (\texttt{HODLR})
algorithm can make the storage and operations both scale in proportion to $N_{time}$, although the 
solver is only approximate \citep{Ambikasaran2013,Ambikasaran2016}.  In addition, the
solver is complex and can be challenging to use with general choices of kernel.  Another
promising approximate technique is to use a uniform resampling of the dataset, and then
rely on the structure of Toeplitz matrices, which require uniformly space time series,
to obtain approximate solutions of the likelihood function \citep{WilsonNickisch2015},
the `Kernel Interpolation for Scalable Structured Gaussian Processes' (\texttt{KISS-GP});
this approach has the drawback of being approximate.  A third approach is to use a kernel
based upon stochastic differential equations:  the `continuous auto-regressive moving-average'
model \citep[aka \texttt{CARMA};][]{Kelly2014}.  This approach can model a stationary power-spectrum that
is the combination of a number of Lorentzians, but has the disadvantage that the amplitudes
of the Lorentzians are required to have a specific relation.  In addition, these
techniques require a stationary kernel. {\color{red} Is this true for GEORGE?}

The approach we explore in this paper is based upon a method developed by Press \&
Rybicki twenty years ago \citep{1995PhRvL..74.1060R}.  They observed that a correlation
matrix consisting of a single exponential kernel has an inverse that is tri-diagonal.
The coefficients and solution of a tri-diagonal matrix scale in proportion to $O(N_{time})$.
They observed how a combination of two exponential kernels would have a simliar scaling,
but they did not demonstrate how to extend this to more than two kernel components.

A significant advance was made by generalizing the Press-Rybicki method to an
arbitrary number of exponential kernels by utilizing the fact that the elements of
such matrices can be written in terms of the products of components of two vectors,
which are so-called semi-separable matrices \citep{Ambikasaran2015}.  This `Generalized
Rybicki-Press' (\texttt{GenRP}) approach
has many advantages:  the computation time and storage both scale as $O(N_{time})$,
the coefficients of the kernels need not be stationary, the computation of the
likelihood is exact, and the algorithm is simple to implement, and thus is robust.
Unlike the \texttt{CARMA} model, there is no restriction on the relation between
the coefficients (except that the kernel needs to be positive definite).  Unlike
\texttt{HODLR} and \texttt{KISS-GP} the method is exact.  The main drawback of
this approach is that the sum of exponential kernels (with real coefficients) can 
only represent a simplified range power spectra of auto-correlated time series
that can be expressed the sum of Lorentzians that peak at zero frequency.

However, this drawback turns out to not be significant:  the exponential kernel
can have a complex exponential coefficients (plus their complex-conjugates), and 
thus approximate an arbitrary autocorrelation function with enough terms, including
quasi-periodic kernels.  It turns out that the sum of Lorentzian components is
a {\it very} good approximation of stellar variability spectra:  the granulation
noise due to stellar convection may be expressed in terms of zero-frequency
Lorentzians, the asteroseismic oscillations may be expressed as non-zero frequency
Lorentizians, as can stellar rotation.  In this paper we extend the \texttt{GenRP} 
approach to kernels with complex exponentials, which when combined with the
complex conjugate leads to damped harmonic auto-correlation functions which
are flexible enough to describe a wide range of variability, and may still
be solved in $O(N_{time}$ operations.

@@

The functional form of $K_j$ may be used to approximate some commonly used Kernel components: radial
kernels can be approximated with $\beta_I=0$, while periodic kernels can be approximated by
$\beta_I \ne 0$.  The squared exponential (Gaussian) kernel and the Mat\'ern kernels are commonly
used radial kernels, while the cosine kernel and the exponential-sine-squared kernels are
commonly used periodic kernels.  Linear combinations of our basis of damped cosine kernels may be 
used to approximate each of these kernels, with various degrees of success:
\begin{itemize}
\item Constant kernel
\item The squared exponential kernel has the behavior of declining exponentially at large
separation, and has $c_1$ = 0.  An infinite number of exponential kernels may be used to
approximate a Gaussian kernel.  For example, to 
\item {\bf Mat\'ern 3/2 kernel.}  The Mat\'ern $3/2$ kernel is given by: 
\begin{equation}
K_{3/2}(\tau,\tau_0)
= (1+x)e^{-x},
\end{equation} 
where $x = \sqrt{3}\vert \tau/\tau_0\vert$.  This may be
well approximated by the sum of two exponential kernels: 
\begin{equation}
K_{3/2}(\tau,\tau_0) \approx (1-b)e^{-x}+be^{-x\frac{b-1}{b}},
\end{equation} 
where $b \gg 1$ is a dimensionless parameter; this approximation is exact in the limit $b \rightarrow \infty$.  For
large values of $b$, the error on this scales as $2/(e^2b) \approx 0.27/b$; for
$b=10$, the error in this approximation is $<3$\%.  Note that the two exponential terms
have opposite signs, which may lead to numerical instability for values of $b$ that
are too large.  One advantage of the Mat\'ern 3/2 kernel is that it is smooth across
$\tau=0$ since its derivative is zero at $\tau=0$.  The exponential kernel {\it does not}
have this property, which leads to qualitatively different noise properties on small
timescales due to the sharpness of the kernel at $\tau=0$.  The Mat\'ern 3/2 kernel
turns out to be an excellent description of stellar granulation, and thus the approximate
formula is simultaneously physically relevant and computationally convenient.
\item Matern 5/2 kernel.
\item Cosine kernel
\item exp-sin-sq kernel
\end{itemize}

>>>>

Stars are variable, for better or worse:  their variability provides information about the properties
of stars, but also results in noise that must be accounted for in detecting and characterizing their
planetary systems.  When analyzing astronomical times series of stars, then, there are generally two 
goals:  1). to analyze the frequency spectrum and amplitude of stellar variability to better understand stellar
properties and evolution, and/or 2). to account for correlated noise when modeling the stellar
flux variability and spectral variations.

A promising technique for addressing both of these goals is the Gaussian Process formalism 
\citep{Rasmussen2006,Gibson2012}.  The basis of this method is to assume that stellar variability
is random, Gaussian noise, but that the random variability is correlated in time, and the nature of the
correlations is completely specified by a correlation matrix.  In its simplest form, the 
correlation matrix is modeled with a time-independent autocorrelation function (although 
the time-independence can be relaxed), including a white-noise component that may vary
with each exposure, and the autocorrelation function is described by a functional form that may be 
simply parameterized by a relatively small number of parameters, sometimes referred to as 
`hyper-parameters.'  The goal of spectral analysis is to measure the parameters which
describe the autocorrelation function, thus inferring properties of the stellar variabilty
which may be used to characterize properties of the star, such as the rotation period,
the asteroseismic frequencies, or granulation noise known as `flicker' \citep{Aerts2010,Noyes1984,Bastien2013}.
The goals of planet detection include measuring the Doppler shift of the star, measuring
the decrement of flux as a planet transits a star, and/or measuring the times of transit to
look for dynamical interactions between planets.  To properly account for the impact of 
stellar variability on the planet parameters requires treating correlations in the data, and 
to do so also requires evaluating the correlation matrix.

The size of the correlation matrix, $N_{elements}$, scales as the square of the length of the time-series,
$N_{element}=N_{time}^2$, so as time series grow in size, say $N_{time} =10^{5-6}$, the storage may become 
prohibitive, $N_{element} = 10^{10-12}$.  In addition, this matrix must be inverted and have
its determinant evaluated to compute the likelihood function, which takes of order $N_{operations}
= N_{time}^3 = 10^{15-18}$.  This becomes an impossible computational problem, which prohibits
using long time series datsets, such as the {\it Kepler} dataset $N_{time} \approx 10^{5-6}$,
{\it Spitzer} time series with $N_{time} \approx 10^5$, and Solar time series with
$N_{time} \approx 10^7$.  So a faster means of solving for the likelihood is required in these
cases.

One promising solution is to use approximate techiques based upon the fact that correlation
matrices display a great deal of symmetry.  This `hierarchical off-diagonal low-rank' (\texttt{HODLR})
algorithm can make the storage and operations both scale in proportion to $N_{time}$, although the 
solver is only approximate \citep{Ambikasaran2013,Ambikasaran2016}.  In addition, the
solver is complicated and can be challenging to use with general choices of kernel.  Another
promising approximate technique is to use a uniform resampling of the dataset, and then
rely on the structure of Toeplitz matrices, which require uniformly space time series,
to obtain approximate solutions of the likelihood function \citep{WilsonNickisch2015},
the `Kernel Interpolation for Scalable Structured Gaussian Processes' (\texttt{KISS-GP});
this approach has the drawback of being approximate.  A third approach is to use a kernel
based upon stochastic differential equations:  the `continuous auto-regressive moving-average'
model \citep[aka \texttt{CARMA};][]{Kelly2014}.  This approach can model a stationary power-spectrum that
is the combination of a number of Lorentzians, but has the disadvantage that the amplitudes
of the Lorentzians are required to have a specific relation.  In addition, these
techniques require a stationary kernel. {\color{red} Is this true for GEORGE?}

The approach we explore in this paper is based upon a method developed by Press \&
Rybicki twenty years ago \citep{1995PhRvL..74.1060R}.  They observed that a correlation
matrix consisting of a single exponential kernel has an inverse that is tri-diagonal.
The coefficients and solution of a tri-diagonal matrix scale in proportion to $O(N_{time})$.
They also indicated that a combination of two exponential kernels would have a simliar scaling,
but they did not demonstrate how to extend this to more than two kernel components.

A significant advance was made in generalizing the Press-Rybicki method to an
arbitrary number of exponential kernels by utilizing the fact that the elements of
such matrices can be written in terms of the products of components of two vectors,
which are so-called semi-separable matrices \citep{Ambikasaran2015}.  This `Generalized
Rybicki-Press' (\texttt{GenRP}) approach
has many advantages:  the computation time and storage both scale as $O(N_{time})$,
the coefficients of the kernels need not be stationary, the computation of the
likelihood is exact, and the algorithm is simple to implement, and thus is robust.
Unlike the \texttt{CARMA} model, there is no restriction on the relation between
the coefficients (except that the kernel needs to be positive definite).  Unlike
\texttt{HODLR} and \texttt{KISS-GP},  the \texttt{GenRP} method is exact.  The main drawback 
of this approach is that the sum of exponential kernels with real coefficients can 
only represent a limited range of power spectra of auto-correlated time series:
those which can be expressed the sum of Lorentzians that peak at zero frequency.

However, this drawback turns out to not be significant:  the exponential kernel
can also have a complex exponential coefficients (plus their complex-conjugates), and 
thus approximate an almost arbitrary autocorrelation function with enough terms, most
importantly allowing for quasi-periodic kernels.  It turns out that the sum of Lorentzian 
components is a {\it very} good approximation of stellar variability spectra:  the 
activity and granulation noise due to stellar convection may be expressed in terms of zero-frequency
Lorentzians \citep{1985ESASP.235..199H}, the asteroseismic oscillations may be expressed as non-zero frequency
Lorentzians \citep{1990ApJ...364..699A,Gruberbauer2009}, and stellar rotation may be 
parameterized this way as well.  In this paper 
we extend the \texttt{GenRP} approach to kernels with complex exponentials, which when 
combined with the complex conjugate leads to damped harmonic auto-correlation functions 
which are flexible enough to describe a wide range of variability, and may still
be solved in $O(N_{time}$ operations.

@@

The functional form of $K_j$ may be used to approximate some commonly used Kernel components: radial
kernels can be approximated with $\beta_I=0$, while periodic kernels can be approximated by
$\beta_I \ne 0$.  The squared exponential (Gaussian) kernel and the Mat\'ern kernels are commonly
used radial kernels, while the cosine kernel and the exponential-sine-squared kernels are
commonly used periodic kernels.  Linear combinations of our basis of damped cosine kernels may be 
used to approximate each of these kernels, with various degrees of success:
\begin{itemize}
\item Constant kernel
\item The squared exponential kernel has the behavior of declining exponentially at large
separation, and has $c_1$ = 0.  An infinite number of exponential kernels may be used to
approximate a Gaussian kernel.  For example, to

\item {\bf Mat\'ern 3/2 kernel.}  The Mat\'ern $3/2$ kernel is given by: 
\begin{equation}
K_{3/2}(\tau,\tau_0) = (1+x)e^{-x},
\end{equation} 
where $x = \sqrt{3}\vert \tau/\tau_0\vert$.  This kernel has a Fourier Transform\footnote{Using the normalization convention: 
$\hat f(\omega) = \frac{1}{\sqrt{2\pi}} \int_{-\infty}^\infty f(t) e^{i\omega t} dt$}  of
\begin{equation}
\hat K_{3/2}(\omega) = \frac{\tau_0 6^{3/2}}{\pi^{1/2}} (3+\tau_0^2\omega^2)^{-2},
\end{equation}
where $\omega = 2\pi f$.  This kernel may be well approximated by the sum of two exponential kernels: 
\begin{equation}
K_{3/2}(\tau,\tau_0) \approx (1-b)e^{-x}+be^{-x\frac{b-1}{b}},
\end{equation} 
where $b \gg 1$ is a dimensionless parameter; this approximation is exact in the limit $b \rightarrow \infty$.  For
large values of $b$, the maximum error on this scales as $2/(e^2b) \approx 0.27/b$; for
$b=10$, the error in this approximation is $<3$\%.  Note that the two exponential terms
have opposite signs, which may lead to numerical instability for values of $b$ that
are large.

One advantage of the Mat\'ern 3/2 kernel is that it is smooth across
$\tau=0$ since its derivative is zero at $\tau=0$.  The exponential kernel {\it does not}
have this property, which leads to qualitatively different noise properties on small
timescales due to the sharpness of the exponential kernel at $\tau=0$.  The Mat\'ern 3/2 kernel
turns out to be an excellent description of stellar activity and granulation, and thus the approximate
formula is simultaneously physically relevant and computationally convenient.  
The power spectrum of a single exponential scales as $f^{-2}$ at high frequency,
while the Mat\'ern kernel scales as $f^{-4}$ which more closely matches the power
spectrum due to activity and granulation in the Sun \citep{2004A&A...414.1139A,2009A&A...495..979M}.
%\item Matern 5/2 kernel.
\item Cosine kernel
\item exp-sin-square kernel
\end{itemize}
